---
title: "Getting started with qio"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(qio)
```

A Parquet file is columnar, compressed, and self-describing: it carries its own
schema and per-chunk statistics in a footer. qio is built around that shape.
Reading a whole file is one call, but the interesting part is that you can read
the footer alone, decide what you need, and decompress only that.

## A whole file at a time

`write_parquet()` and `read_parquet()` are the eager pair.

```{r}
path <- tempfile(fileext = ".parquet")
write_parquet(mtcars, path)

cars <- read_parquet(path)
dim(cars)
```

Row names are not a Parquet concept, so they do not survive the round trip.
Everything else does, subject to the type mapping in `?qio-types`.

## Looking before reading

The rest of this article uses a bundled file written by pyarrow, so the
inspection functions have something real to report.

```{r}
path <- system.file("extdata", "bloom_sorted.parquet", package = "qio")
pf <- open_parquet(path)
pf
```

`open_parquet()` reads the footer, not the data. The handle is cheap, and
everything below answers from metadata alone.

```{r}
schema(pf)
```

`schema()` reports what the *file* says. `read_plan()` reports what R will
actually get, which is the question you usually have:

```{r}
read_plan(pf)[, c("path", "physical_type", "logical_type", "r_type")]
```

The plan is a pure function of the schema and your read options, so it is the
same for `read_parquet()`, `collect()`, and `walk_batches()`. If a column would
arrive as something you did not expect, you find out here rather than after a
long read.

Row groups are the unit of selective reading:

```{r}
row_groups(pf)
```

## Reading only what you need

A column you do not select is never decompressed, which is the cheapest
speed-up available on a wide file.

```{r}
head(collect(pf, columns = c("key", "label")), 3)
```

Row groups select the other axis, and the two compose:

```{r}
nrow(collect(pf, row_groups = 1))
nrow(collect(pf, columns = "key", row_groups = c(1, 2)))
```

## Reading from a URL

`read_parquet()` and `open_parquet()` accept an `http://`, `https://`, `ftp://`,
`ftps://` or `file://` URL as well as a path:

```{r, eval = FALSE}
remote <- read_parquet("https://example.com/data.parquet")
```

The whole file is downloaded to the session temporary directory before any of it
is read, and the copy is removed when the read finishes -- or, for a handle,
when you call `close_parquet()`. Selecting columns or row groups from a URL
therefore saves decoding but not transfer, so for a large remote file it is
usually better to download once yourself and read the local copy repeatedly.
`?qio-limitations` explains why qio cannot fetch only the parts it needs.

Writing to a URL is refused rather than silently written somewhere local.

## Files larger than memory

`walk_batches()` calls your function once per batch and keeps one batch alive
at a time, so peak memory follows `batch_size` rather than file size.

```{r}
total <- 0
walk_batches(pf, batch_size = 1000, FUN = function(batch, index) {
  total <<- total + sum(batch$score)
})
total
```

## What the file knows about itself

Parquet stores per-chunk statistics, and writers may add a page index or a
bloom filter. qio surfaces all three, which is how you prune work before doing
it.

```{r}
column_statistics(pf)[1:3, c("row_group", "path", "null_count", "min", "max")]
```

A bloom filter answers exactly one question -- is this value *definitely
absent* -- and never proves presence:

```{r}
bloom_filter_may_contain(pf, "key", c(42, 123456))
```

`FALSE` is a guarantee; `TRUE` means "maybe, go and read". qio's own writer
emits no bloom filters, which is why this file came from elsewhere.

```{r}
close_parquet(pf)
```

Handles hold an open file. Close them when you are done; `read_parquet()` opens
and closes one for you.

## Types worth knowing about

Two mappings surprise people, and both are deliberate.

**64-bit integers.** R has no native 64-bit integer. By default `INT64` becomes
`double`, which is exact to 2^53 and `NA` beyond it, with one warning naming the
column. Pass `int64 = "integer64"` for the full range, which needs the suggested
`bit64` package.

**Bytes are not text.** An unannotated `BYTE_ARRAY` is arbitrary bytes and reads
as a list of raw vectors. It becomes `character` only when the file says so with
a `STRING`, `ENUM`, or `JSON` annotation. Text is validated as UTF-8, and an
invalid value fails with its column and row rather than arriving as mojibake.

`?qio-types` documents every mapping and where precision is lost.

## Writing

Inference covers the ordinary R types. Where one R type could mean several
Parquet types, `parquet_schema()` decides:

```{r}
types <- parquet_schema(mpg = "FLOAT", cyl = "INT64")
out <- tempfile(fileext = ".parquet")
write_parquet(mtcars, out, schema = types, compression = "zstd")

infer_parquet_schema(mtcars[, c("mpg", "cyl")])
```

`write_parquet()` also takes `row_group_size`, `metadata`, `sorted_by`, and
`append`.

## What qio does not do

Nested and repeated columns are skipped with a message, decimals read as
`double` rather than exact fixed-point, and several types can be read but not
written. None of that is accidental; `?qio-limitations` records each one and
why, so you can tell a deliberate boundary from a bug.
